Appendix: Why the complement of AUC isn't a casewise CV criterion

### 2024-07-04 scratch/notes_georges/JSS-article/JSS-article-3-GM.Rmd
reduce example 3.2 to 7 3/4 pages

Data_template <- expand.grid(patient = 1:Nb, obs = 1:Nw) |>
  within({
    xw <- seq(-2, 2, length.out = Nw)[obs]
    x <- patient + xw
    xm  <- ave(x, patient)   # within-patient mean
    # Scaled random error within each SD_ratio_i group
    re_std <- scale(resid(lm(rnorm(Nb*Nw) ~ x)))
    re_between <- ave(re_std, patient)
    re_within <- re_std - re_between
    re_between <- scale(re_between)/sqrt(Nw)
    re_within <- scale(re_within)
  })
Data <- do.call(
  rbind,
  lapply(
    1:Nv,
    function(i) {
      cbind(Data_template, SD_ratio_i = i)
    }
  )
)
Data <- within(
  Data,
  {
    SD_within_ <- SD_within[SD_ratio_i]
    SD_between_ <- SD_between[SD_ratio_i]
    SD_ratio <- SD_ratio[SD_ratio_i]
    y <- 10 +
      Bb * xm +                  # contextual effect
      Bw * (x - xm) +            # within-patient effect
      SD_within_ * re_within +   # within patient random effect
      SD_between_ * re_between   # adjustment to between patient random effect
  }
)
```
Figure \ref{fig:plot1} display scatterplots of each data set showing estimated 50% concentration ellipses for each cluster.[^lattice]

@@

```{r plot1, echo=FALSE}
#| out.width = "100%",
#| fig.height = 3.5,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the between-patient SD/within-patient SD. The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel."
library("lattice")
library("latticeExtra")
layer <- latticeExtra::layer # to avert conflicts with ggplot2::layer
xyplot(y ~ x | SD_ratio, data = Data, group = patient,
       par.strip.text = list(cex = 0.7),
       scales = list(x = list(alternating = FALSE)),
       layout = c(Nv, 1),
       par.settings = list(superpose.symbol = list(pch = 1, cex = 0.7))) +
  layer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                      level = 0.5),
        panel.abline(a = 10, b = 1))
```
The population from which these datasets are generated has a between-patient effect of dosage of 1 and a within-patient effect of $-0.5$. The researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models have sequentially nested fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`. The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`. We thus fit four models to four datasets for a total of 16 models.

@@

```{r echo=FALSE,include=FALSE}
library(cv) # unclear why it's necessary to reload cv
.opts <- options(warn = -2) 
```
```{r cross-validation,cache=TRUE,message=FALSE}
model_lists <- lapply(fits, function(fitlist) do.call(models, fitlist))
cvs_cases <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]], k = 10, 
              data = split(Data, ~ SD_ratio)[[i]])
         })
cvs_clusters <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]],  k = 10, 
              data = split(Data, ~SD_ratio)[[i]], 
              clusterVariables = 'patient')
         })
```
For a given dataset, we can plot the results for a list of models using the plot method for `"cvList"` objects. For example, this code produces Figure \ref{fig:plot-cv-example}.
```{r plot-cv-example}
#| out.width = "60%",
#| fig.height = 3.5,
#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept mixed models with varying fixed effects."
plot(cvs_clusters[[2]], main="Comparison of Fixed Effects",)
```
We assemble the results for all datasets to show them in a common figure.
```{r cross-validation-data}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))
dsummary$cases <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
  })
dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r cross-validation-data-plot}
#| out.width = "100%",
#| fig.height = 4,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(cases + clusters ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', reverse.rows = T, columns = 2), 
       type = 'b', ylab = 'CV criterion (MSE)',
       par.strip.text = list(cex = 0.7),
       layout= c(Nv, 1),
       par.settings =
         list(superpose.line=list(lty = c(1, 2), lwd = 2),
              superpose.symbol=list(pch = 15:16, cex = 1)),
       scales = list(y = list(log = TRUE), 
                     x = list(alternating = F, rot = 45))) 
```
In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. With all our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

>>>>

Data_template <- expand.grid(patient = 1:Nb, obs = 1:Nw) |>
  within({
    xw <- seq(-2, 2, length.out = Nw)[obs]
    x <- patient + xw
    xm  <- ave(x, patient)   # within-patient mean
    # Scaled random error within each SD_ratio_i group
    re_std <- scale(resid(lm(rnorm(Nb*Nw) ~ x)))
    re_between <- ave(re_std, patient)
    re_within <- re_std - re_between
    re_between <- scale(re_between)/sqrt(Nw)
    re_within <- scale(re_within)
  })
Data <- do.call(
  rbind,
  lapply(
    1:Nv,
    function(i) {
      cbind(Data_template, SD_ratio_i = i)
    }
  )
)
Data <- within(
  Data,
  {
    SD_within_ <- SD_within[SD_ratio_i]
    SD_between_ <- SD_between[SD_ratio_i]
    SD_ratio <- SD_ratio[SD_ratio_i]
    y <- 10 +
      Bb * xm +                  # contextual effect
      Bw * (x - xm) +            # within-patient effect
      SD_within_ * re_within +   # within patient random effect
      SD_between_ * re_between   # adjustment to between patient random effect
  }
)
```
Figure \ref{fig:plot1} displays scatterplots of each data set showing estimated 50% concentration ellipses for each cluster.[^lattice]

@@

```{r plot1, echo=FALSE}
#| out.width = "100%",
#| fig.height = 3.5,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the between-patient SD/within-patient SD. The ellipses are estimated 50% concentration ellipses for each patient. The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel."
library("lattice")
library("latticeExtra")
layer <- latticeExtra::layer # to avert conflicts with ggplot2::layer
xyplot(y ~ x | SD_ratio, data = Data, group = patient,
       par.strip.text = list(cex = 0.7),
       scales = list(x = list(alternating = FALSE)),
       layout = c(Nv, 1),
       par.settings = list(superpose.symbol = list(pch = 1, cex = 0.7))) +
  layer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                      level = 0.5),
        panel.abline(a = 10, b = 1))
```
The population from which these datasets are generated has a between-patient effect of dosage of 1 and a within-patient effect of $-0.5$. The researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models have sequentially nested fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`. The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`. We thus fit four models to four datasets for a total of 16 models.

@@

```{r echo=FALSE,include=FALSE}
library(cv) # unclear why it's necessary to reload cv
.opts <- options(warn = -2) 
```
```{r cross-validation,cache=TRUE,message=FALSE}
model_lists <- lapply(fits, function(fitlist) do.call(models, fitlist))
cvs_cases <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]], k = 10, 
              data = split(Data, ~ SD_ratio)[[i]])
         })
cvs_clusters <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]],  k = 10, 
              data = split(Data, ~SD_ratio)[[i]], 
              clusterVariables = 'patient')
         })
```
For a given dataset, we can plot the results for a list of models using the plot method for `"cvList"` objects. For example, this code produces Figure \ref{fig:plot-cv-example}.
```{r plot-cv-example}
#| out.width = "60%",
#| fig.height = 3.5,
#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept mixed models with varying fixed effects."
plot(cvs_clusters[[2]], main="Comparison of Fixed Effects",)
```
The results for all datasets are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.
```{r cross-validation-data,include=FALSE}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))
dsummary$cases <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
  })
dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r cross-validation-data-plot,echo=FALSE}
#| out.width = "100%",
#| fig.height = 3.5,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(cases + clusters ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', reverse.rows = T, columns = 2), 
       type = 'b', ylab = 'CV criterion (MSE)',
       par.strip.text = list(cex = 0.7),
       layout= c(Nv, 1),
       par.settings =
         list(superpose.line=list(lty = c(1, 2), lwd = 2),
              superpose.symbol=list(pch = 15:16, cex = 1)),
       scales = list(y = list(log = TRUE), 
                     x = list(alternating = F, rot = 45))) 
```
In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. With all our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.
